Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-10-06 psy_peds_3.Rmd
table 1 + updating plots

```{r}
# cut-off dates used in the analysis, these should not be changed
history_start_date <- "2019-01-01"
history_end_date <- "2021-05-31"
pandemic_start_date <- "2020-03-15"
start_date_plots <- "2019-01-15"
end_date_plots <- "2021-05-01"
```

@@

```{r}
# Model without seasonality
time_series_analysis_plotting <- function(model, df_count_psy, count_col, clearance_period = FALSE) {
  output_model <- broom::tidy(model)
  coeff_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "estimate"][[1]], 3)
  pvalue_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "p.value"][[1]], 3)
  annotation <- paste0("Lockdown slope: ", coeff_slope, " (pvalue: ", pvalue_slope, ")")
  pred_model <- predict(model, patient_count_monthly_psy, type = "response", interval = "confidence") %>% as.data.frame()
  pred_model_b <- predict(model, mutate(patient_count_monthly_psy, period = "before_pandemic"), type = "response", interval = "confidence") %>%
    as.data.frame()
  names(pred_model_b) <- paste0(names(pred_model_b), "_wo_intervention")
  pred_model <- bind_cols(df_count_psy, pred_model, pred_model_b)
  if (clearance_period) {
    pred_model$clearance_period <- ifelse(pred_model$time_p < as.Date(cut(start_clear_period, breaks = time_period)) ,
                                          "before_pandemic",
                                          ifelse(pred_model$time_p > as.Date(cut(end_clear_period, breaks = time_period)),
                                                 "during_pandemic",
                                                 "during_clearance")
    )
    pred_model$lwr <- ifelse(pred_model$clearance_period == "during_clearance", NA, pred_model$lwr)
    pred_model$upr <- ifelse(pred_model$clearance_period == "during_clearance", NA, pred_model$upr)
    pred_model$fit_wo_intervention <- ifelse(pred_model$time_p >= max(pred_model$time_p[pred_model$clearance_period == "before pandemic"]),
                                             pred_model$fit_wo_intervention,
                                             NA)
  } else {
    pred_model$clearance_period <- pred_model$period
  }
  print(pred_model$clearance_period)
  plot <- ggplot(pred_model, aes_string(x = "time_p", y = count_col)) +
    geom_point(aes(colour = clearance_period)) +
    geom_line(aes(x = time_p, y = fit, colour = clearance_period)) +
    geom_line(aes(x = time_p, y = fit_wo_intervention, colour = "Expected mean w/o pandemic"), linetype = "dashed") +
    geom_ribbon(aes(ymin = lwr, ymax = upr, colour = clearance_period), alpha = 0.3, colour = NA) +
    scale_color_manual(name = "Time Series projection", values = c("before_pandemic" = "black",
                                                                   "Expected mean w/o pandemic" = "red",
                                                                   "during_pandemic" = "black",
                                                                   "during_clearance" = "white")) +
    annotate("text", x = as.Date("2019-10-01"), y = max(pred_model[[count_col]]) + 5, label = annotation) +
    theme(legend.position = "bottom")

@@

```{r}
# Time series analysis of percentage of psy codes per encounter, linear (without seasonality)
patient_count_monthly_psy <- count_icd_psy %>%
  filter(time_p > as.Date(cut( start_date_plots, breaks = time_period)) &
           time_p < as.Date(cut( end_date_plots, breaks = time_period))) %>%
  mutate(time = seq_along(time_p),
         month_year = format(time_p, "%m")
  )
start_clear_period <- as.Date("2020-03-15")
end_clear_period <- as.Date("2020-05-15")
patient_count_monthly_psy_clear <- patient_count_monthly_psy %>%
  mutate( count_psy = ifelse(time_p < as.Date(cut( start_clear_period, breaks = time_period)) |
                               time_p > as.Date(cut( end_clear_period, breaks = time_period)),
                             count_psy,
                             NA)
  )
```

@@

>>>>

```{r}
# cut-off dates used in the analysis, these should not be changed
history_start_date <- as.Date("2019-01-01")
history_end_date <- as.Date("2021-05-31")
pandemic_start_date <- as.Date("2020-03-15")
start_date_plots <- as.Date("2019-01-15")
end_date_plots <- as.Date("2021-05-01")
```

@@

```{r}
# Model without seasonality
time_series_analysis_plotting <- function(model, df_count_psy, count_col, clearance_period = FALSE) {
  output_model <- broom::tidy(model)
  coeff_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "estimate"][[1]], 3)
  pvalue_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "p.value"][[1]], 3)
  annotation <- paste0("Lockdown slope: ", coeff_slope, " (pvalue: ", pvalue_slope, ")")
  pred_model <- predict(model, patient_count_monthly_psy, type = "response", interval = "confidence") %>% as.data.frame()
  pred_model_b <- predict(model, mutate(patient_count_monthly_psy, period = "before_pandemic"), type = "response", interval = "confidence") %>%
    as.data.frame()
  names(pred_model_b) <- paste0(names(pred_model_b), "_wo_intervention")
  pred_model <- bind_cols(df_count_psy, pred_model, pred_model_b)
  if (clearance_period) {
    pred_model$clearance_period <- ifelse(pred_model$time_p < as.Date(cut(start_clear_period, breaks = time_period)) ,
                                          "before_pandemic",
                                          ifelse(pred_model$time_p > as.Date(cut(end_clear_period, breaks = time_period)),
                                                 "during_pandemic",
                                                 "during_clearance")
    )
    pred_model$lwr <- ifelse(pred_model$clearance_period == "during_clearance", NA, pred_model$lwr)
    pred_model$upr <- ifelse(pred_model$clearance_period == "during_clearance", NA, pred_model$upr)
    pred_model$fit_wo_intervention <- ifelse(pred_model$time_p >= max(pred_model$time_p[pred_model$clearance_period == "before pandemic"]),
                                             pred_model$fit_wo_intervention,
                                             NA)
  } else {
    pred_model$clearance_period <- pred_model$period
  }
  print(pred_model$clearance_period)
  plot <- ggplot(pred_model, aes_string(x = "time_p", y = count_col)) +
    geom_point(aes(colour = clearance_period)) +
    geom_line(aes(x = time_p, y = fit_wo_intervention, colour = "Expected mean w/o pandemic"), linetype = "dashed") +
    geom_line(aes(x = time_p, y = fit, colour = clearance_period)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr, colour = clearance_period), alpha = 0.3, colour = NA) +
    scale_color_manual(name = "Time Series projection", values = c("before_pandemic" = "black",
                                                                   "Expected mean w/o pandemic" = "red",
                                                                   "during_pandemic" = "black",
                                                                   "during_clearance" = "white")) +
    annotate("text", x = as.Date("2019-10-01"), y = max(pred_model[[count_col]]) + 5, label = annotation) +
    theme(legend.position = "bottom")

@@

```{r}
# Time series analysis of percentage of psy codes per encounter, linear (without seasonality)
patient_count_monthly_psy <- count_icd_psy %>%
  filter(time_p > as.Date(cut( start_date_plots, breaks = time_period)) &
           time_p < as.Date(cut( end_date_plots, breaks = time_period))) %>%
  mutate(time = seq_along(time_p),
         month_year = format(time_p, "%m")
  )

start_clear_period <- as.Date("2020-03-15")
end_clear_period <- as.Date("2020-05-15")
patient_count_monthly_psy_clear <- patient_count_monthly_psy %>%
  mutate( count_psy = ifelse(time_p < as.Date(cut( start_clear_period, breaks = time_period)) |
                               time_p > as.Date(cut( end_clear_period, breaks = time_period)),
                             count_psy,
                             NA)
  )
```

@@

## Table 1

```{r}

t.test2 <- function(m1, m2, s1, s2, n1, n2) {
  # t.test computed from sample statistics, ie using mean, standard deviation and count value of the samples
  # useful in case we can only retrieve aggregated stats from the sites
  # ref: https://stats.stackexchange.com/questions/30394/how-to-perform-two-sample-t-tests-in-r-by-inputting-sample-statistics-rather-tha
  # and https://en.wikipedia.org/wiki/Student%27s_t-test?oldformat=true#Equal_or_unequal_sample_sizes,_unequal_variances_(sX1_%3E_2sX2_or_sX2_%3E_2sX1)
  se <- sqrt( (s1^2/n1) + (s2^2/n2) )
  # welch-satterthwaite df
  df <- ( (s1^2/n1 + s2^2/n2)^2 )/( (s1^2/n1)^2/(n1-1) + (s2^2/n2)^2/(n2-1) )
  t <- (m1-m2)/se
  dat <- c(m1-m2, se, t, 2*pt(-abs(t),df))
  names(dat) <- c("Difference of means", "Std Error", "t", "p-value")
  return(dat)
}

age_by_period <- all_codes_psy %>%
  distinct(patient_num, age_time_diagnosis, period) %>% 
  dplyr::group_by(period) %>%
  dplyr::summarise( mean = mean( age_time_diagnosis ),
                    sd = sd( age_time_diagnosis ),
                    count = n_distinct(patient_num))

age_test <- t.test2(m1 = age_by_period[[1, "mean"]],
                    m2 = age_by_period[[2, "mean"]],
                    s1 = age_by_period[[1, "sd"]],
                    s2 = age_by_period[[2, "sd"]],
                    n1 = age_by_period[[1, "count"]],
                    n2 = age_by_period[[2, "count"]]) %>%
  `[[`("p-value")
```

## Sex
```{r}
sex <- all_codes_psy %>%
  dplyr::distinct(patient_num, sex) %>% 
  dplyr::group_by( sex ) %>%
  dplyr::summarise( count = n_distinct(patient_num) ) %>%
  dplyr::mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5),
                 total = sum( count ), percentage = round( count/total*100, 2)  )

sex_by_period <- all_codes_psy %>%
  dplyr::distinct(patient_num, period, sex) %>% 
  group_by(period, sex ) %>%
  summarise(count = n_distinct(patient_num)) %>%
  group_by(period, .add=TRUE) %>%
  mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5),
          percentage=round(count/sum( count )*100, 2)) %>%
  replace_na(list(male = 0, female = 0))

sex_test <- sex_by_period %>%
  select(-percentage) %>%
  pivot_wider(names_from = sex, values_from = count) %>%
  ungroup() %>%
  select(male, female) %>%
  chisq.test() %>%
  `[[`("p.value")
```

## Diagnosis

### By ICD code
```{r}
summary_diagnosis <- all_codes_psy %>%
  dplyr::distinct( patient_num, description, period, weeks, concept_code ) %>%
  dplyr::group_by( concept_code, description, period ) %>%
  dplyr::count( patient_num ) %>%
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc(count) ) %>%
  dplyr::ungroup()

summary_diagnosis_test <- summary_diagnosis %>%
  tidyr::pivot_wider(names_from = period,
                     values_from = count) %>%
  #select(late_pandemic, early_pandemic) %>%
  select(during_pandemic) %>%
  #tidyr::replace_na(list(late_pandemic = 0, early_pandemic = 0)) %>%
  tidyr::replace_na(list(during_pandemic = 0)) %>%
  chisq.test() %>%
  `[[`("p.value")

### By disorder group
```{r}
summary_disorder_group <- all_codes_psy %>%
  dplyr::group_by( disorder_group ) %>%
  dplyr::count( patient_num ) %>%
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc( count ))

summary_disorder_group_by_period <- all_codes_psy %>%
  dplyr::distinct( patient_num, disorder_group, period, weeks ) %>%
  dplyr::group_by( period, disorder_group ) %>%
  summarise(count = n_distinct(patient_num)) %>%
  group_by(period, add=TRUE) %>%
  mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5),
         percentage = count/sum( count )*100)

summary_disorder_group_by_period_test <- summary_disorder_group_by_period %>%
  pivot_wider(id_cols = disorder_group, names_from = period, values_from = c(count, percentage)) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, replace = 0))) %>%
  #select(count_early_pandemic, count_late_pandemic) %>%
  select(count_during_pandemic) %>%
  chisq.test() %>%
  `[[`("p.value")

```{r}
format_counts <- function(count, percentage) {
  glue::glue("{count} ({percentage} %)", count = count, percentage = percentage)
}

age_table1 <- age_by_period %>% 
  mutate(result = glue::glue("{mean} (sd: {sd})", mean = round(mean, 2), sd = round(sd, 2))) %>% 
  select(result, period) %>% 
  pivot_wider( names_from = period, values_from = result ) %>%
  mutate(name = "Age")

sex_table1 <- sex_by_period %>%
  bind_rows(
    sex_by_period %>%
      group_by(period) %>%
      dplyr::summarise(sex = "Total", count = sum(count), percentage = sum(percentage))
  ) %>%
  mutate(result = format_counts(count, percentage)) %>%
  select(-c(count, percentage)) %>%
  pivot_wider( names_from = period, values_from = result ) %>%
  rename(name = sex)

disorder_table1 <- summary_disorder_group_by_period %>%
  arrange(desc(count)) %>% 
  bind_rows(
    dplyr::summarise(., disorder_group="Total", across(where(is.numeric), sum))
  ) %>%
  mutate(percentage = round(percentage, 2)) %>% 
  mutate(result = format_counts(count, percentage)) %>%
  select(-c(count, percentage)) %>%
  pivot_wider( names_from = period, values_from = result ) %>%
  replace_na( list(during_pandemic = "0 (0 %)")) %>%
  rename(name = disorder_group)
```

```{r}
list_tests <- c("Sex" = sex_test,
                "Age" = age_test,
                "Diagnoses" = summary_diagnosis_test,
                "Disorder groups" = summary_disorder_group_by_period_test
)

list_table1 <- list(
  "Sex" = sex_table1,
  "Age" = age_table1,
  "Disorder groups" = disorder_table1)

tests <- list_tests %>%
  data.frame() %>%
  round(2)
names(tests) <- "pvalue"
tests["Category"] <- row.names(tests)

table1 <- list_table1 %>%
  lapply(function(df) mutate(df, across(.fns = as.character))) %>%
  dplyr::bind_rows(.id = "Category") %>%
  dplyr::left_join(tests, by = "Category")

table1 %>%
  group_by(Category) %>%
  mutate(pvalue = ifelse(duplicated(pvalue, fromLast = TRUE), "", as.character(pvalue))) %>%
  ungroup() %>%
  mutate(Category = "") %>%
  mutate(across(.cols = c(before_pandemic, during_pandemic), .fns = replace_na, "0")) %>%
  rename_with(str_to_title) %>%
  rename_with(sub, pattern = "_", replacement = " ") %>%
  mutate(across(.fns = str_to_title)) %>%
  mutate(across(.fns = sub, pattern = "_", replacement = " ")) %>%
  kable( booktabs = T) %>%
  kableExtra::kable_styling(latex_options = "striped") %>%
  kableExtra::group_rows(index = setNames(rle(table1$Category)[[1]], rle(table1$Category)[[2]]))
```
